---
title: "Getting Started with SIMPLE: Platform Trial Simulation"
output: 
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Getting Started with SIMPLE}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

```{r setup, message = FALSE}
suppressPackageStartupMessages(library(simple))
```

# Introduction

## What are Platform Trials?

**Platform trials** are innovative, adaptive clinical trials that allow multiple investigational treatments to be evaluated simultaneously against a common control arm within a single trial infrastructure. Unlike traditional two-arm trials, platform trials:

- **Add treatments over time**: New investigational arms (ISAs - Intervention-Specific Appendices) can enter the platform after it has started
- **Drop ineffective treatments**: Arms showing futility can be stopped early while the platform continues
- **Share control data**: A common control arm can be shared across multiple treatment comparisons, improving efficiency
- **Conduct interim analyses**: Regular decision points allow for adaptive modifications
- **Continue indefinitely**: The platform remains open as long as there are research questions to answer

This design is particularly valuable in areas where multiple therapies need evaluation (e.g., oncology, infectious diseases, rare diseases) and can significantly reduce the time and cost of drug development.

## Why SIMPLE?

**SIMPLE** (**SIM**ulating **PL**atform trials **E**fficiently) is an R package designed to help researchers, statisticians, and sponsors design and evaluate platform trials through simulation. 

### Key Features

- **Modular architecture**: Each aspect of the trial (recruitment, analysis, decision rules) is handled by independent, reusable modules
- **Out-of-the-box designs**: Pre-built functions for common platform trial designs let you start simulating immediately
- **Flexible customization**: Advanced users can modify individual modules or create entirely new trial designs
- **Operating characteristics**: Built-in functions to run thousands of simulations and calculate power, type I error, sample size, and duration
- **Parallel computing**: Efficiently run simulations across multiple CPU cores

# Quick Start: Your First Platform Trial Simulation

Let's simulate a simple platform trial with 3 treatment arms and a shared control:

```{r quickstart, eval = FALSE}
# Step 1: Define the trial design (using defaults)
trial_design <- fnSimpleDesign()

# Step 2: Simulate a single trial
single_trial <- fnRunSingleTrialSim(trial_design)

# Step 3: View results
single_trial$lSummary
```

This creates a platform trial where treatments enter over time, are analyzed at interim looks, and decisions are made about efficacy/futility.

# Understanding the Modular Architecture

SIMPLE organizes platform trial simulation into **three types of modules**:

## 1. Platform-Level Modules

These modules govern aspects that apply to the entire platform:

| Module | Purpose | Example |
|--------|---------|---------|
| **`lNewIntr`** | Controls when/how new ISAs enter | Add 1 ISA every 6 months until 5 total |
| **`lAddIntr`** | Handles technical setup of new ISAs | Initialize data structures for new arm |
| **`lRecrPars`** | Defines patient recruitment | Recruit 20 patients per week |
| **`lAllocIntr`** | Assigns patients to ISAs | Proportional to number of open arms |
| **`lSimBase`** | Simulates baseline covariates | Age ~Uniform(50,70), 55% male |
| **`lStopRule`** | Determines platform closure | Stop after 1000 patients or 5 years |
| **`lPltfSummary`** | Summarizes trial results | Extract decisions, duration, sample size |

## 2. ISA-Level Modules

These modules can differ between investigational arms:

| Module | Purpose | Example |
|--------|---------|---------|
| **`lAllocArm`** | Assigns patients within ISA | 2:1 randomization (treatment:control) |
| **`lPatOutcome`** | Simulates patient outcomes | Response rate: control=20%, treatment=35% |
| **`lCheckAnlsMstn`** | Triggers analyses | Conduct analysis at 50, 100, 150 patients |
| **`lAnls`** | Performs statistical tests | Chi-squared test for binary endpoint |
| **`lSynthRes`** | Makes decisions from results | Stop for efficacy if p<0.01 |
| **`lCheckEnrl`** | Controls ISA enrollment | Stop enrolling after decision |

## 3. Workflow Functions

These functions tie everything together:

- **`fnRunSingleTrialSim()`**: Simulates one complete platform trial
- **`fnSimDsgnOC()`**: Runs many trials and calculates operating characteristics
- **`fnSimpleDesign()`**, **`fnWP6simple()`**: Pre-built trial designs

# A Realistic Example: NASH Platform Trial

Let's walk through a more realistic example based on a non-alcoholic steatohepatitis (NASH) platform trial design.

## Trial Design Specifications

- **Primary endpoint**: Binary response (NASH resolution OR fibrosis improvement)
- **Sample size**: 100 patients per arm
- **Interim analyses**: At 25%, 50%, 75%, and 100% information
- **Decision rules**: Stop for efficacy if p<0.01, futility if p>0.5
- **Recruitment**: 10 patients per week
- **ISAs**: Start with 2, add new ISA when one completes, maximum 5 total
- **Control arm**: Shared across all ISAs

## Setting Up the Simulation

```{r nash_example, eval = FALSE}
# Create the trial design with custom parameters
nash_design <- fnSimpleDesign(
  nISAs = 5,                    # Maximum number of ISAs
  nPats = 100,                  # Patients per arm
  cGroups = c("Control", "Treatment"),
  dTheta = c(0.20, 0.35),       # Response rates: 20% control, 35% treatment
  nStartIntr = 2,               # Start with 2 ISAs
  analysis_times = c(0.25, 0.5, 0.75, 1.0)  # Interim looks
)

# Simulate a single trial to see what happens
trial_result <- fnRunSingleTrialSim(nash_design)

# Examine when ISAs entered and their outcomes
trial_result$lSummary$Start_Time    # When each ISA started
trial_result$lSummary$Decision      # Final decision for each ISA
trial_result$lSummary$Avg_Time      # Trial duration
```

## Calculating Operating Characteristics

To evaluate the design, we need to simulate many trials:

```{r ocs_example, eval = FALSE}
# Run 1000 simulations across 4 CPU cores
operating_chars <- fnSimDsgnOC(
  lPltfDsgn = nash_design,
  nIter = 1000,
  nCores = 4
)

# View results
operating_chars$Power               # Probability of correctly identifying effective treatment
operating_chars$Type1Error          # False positive rate
operating_chars$Avg_N_Total         # Average total sample size
operating_chars$Avg_Time            # Average trial duration
```

# Customizing Your Design

## Modifying Default Values

The easiest way to customize is by changing function arguments:

```{r custom_args, eval = FALSE}
# Change recruitment pattern
custom_design <- fnSimpleDesign(
  nISAs = 5,
  nPats = 150,                  # Larger sample size
  nStartIntr = 1,               # Start with only 1 ISA
  analysis_times = c(0.5, 1.0)  # Only 1 interim + final
)
```

## Creating Custom Modules

For more control, create your own modules. Example: weekly recruitment that increases over time:

```{r custom_module, eval = FALSE}
# Create custom recruitment module
my_recruitment <- new_lRecrPars(
  fnRecrProc = function(lPltfTrial, lAddArgs) {
    # Ramp up: start with 5 patients/week, increase to 20 after 12 weeks
    current_week <- lPltfTrial$lSnap$dCurrTime
    if (current_week <= 12) {
      return(5)
    } else {
      return(20)
    }
  },
  lAddArgs = list()  # No additional arguments needed
)

# Use it in your design
custom_design <- fnSimpleDesign(nISAs = 5, nPats = 100)
custom_design$lRecrPars <- my_recruitment  # Replace default recruitment
```

# Module Details

Each module has its own detailed vignette with examples:

- **`lNewIntr`**: ISA inclusion timing and rules
- **`lRecrPars`**: Patient recruitment patterns
- **`lAddIntr`**: Adding ISAs to the platform
- **`lAddPats`**: Adding patients to the trial
- **`lAllocIntr`**: Between-ISA allocation
- **`lCheckEnrl`**: Enrollment monitoring
- **`lSnap`**: Trial state snapshots
- **`lStopRule`**: Platform stopping rules

# Next Steps

1. **Explore module vignettes**: Each module has detailed documentation with examples
2. **Try the examples**: Run the code in this vignette and modify parameters
3. **Design your trial**: Start with `fnSimpleDesign()` and customize as needed
4. **Evaluate scenarios**: Use `fnSimDsgnOC()` to compare different design choices
5. **Get help**: Visit https://github.com/el-meyer/simple for documentation and issues

# Key References

- Meyer EL, et al. (2023). SIMPLE—A modular tool for simulating complex platform trials. *SoftwareX*, 23, 101515.
- Angus DC, et al. (2019). Adaptive platform trials: definition, design, conduct and reporting considerations. *Nature Reviews Drug Discovery*, 18, 797-807.




